---
title: "PhosR STAR protocols"
author: "Hani Kim"
date: "09/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Step 1: Installing PhosR
Install PhosR by running the following code:
```{r}
if(!require(devtools)){ 
 install.packages("devtools") # If not already installed 
}

devtools::install_github("PYangLab/PhosR", 
                         build_opts = c("--no-resave-data", "--no-manual"), 
                         build_vignettes = TRUE)

library(PhosR)
```

Whilst this STAR protocol provides an in-depth step-by-step tutorial for running PhosR with tips and suggestions, you can browse through the streamlined version of the vignette by running the following code. 
```{r}
utils::browseVignettes("PhosR")
```

```{r}
suppressPackageStartupMessages(library(SummarizedExperiment))
```

# Step 2: Creating the PhosphoExperiment Object
```{r}
# Load dataset
load("../Data/PXD001792_raw_hepatocyte.RData", verbose = TRUE)

# Extract the columns of interests before start 
samples = colnames(PXD001792_raw_hepatocyte)
PXD001792_raw_hepatocyte <- PXD001792_raw_hepatocyte[,c(grep("_Control_\\d$", samples),
                                                        grep("_Ins_\\d$", samples))]
PXD001792_raw_hepatocyte = PXD001792_raw_hepatocyte[, sort(colnames(PXD001792_raw_hepatocyte))]

# Create PhosphoExperiment Object
ppe <- PhosR::PhosphoExperiment(assays = list(Quantification = as.matrix(PXD001792_raw_hepatocyte)))

# Add site annotations to PhosphoExperiment object
splitSites = strsplit(rownames(ppe), "~")
GeneSymbol <- toupper(as.character(sapply(splitSites, function(x) x[[2]])))
Residue <- as.character(gsub("[0-9]", "", sapply(splitSites, function(x) x[[3]])))
Site <- as.numeric(gsub("[STY]","",sapply(splitSites, function(x) x[[3]])))
Sequence <- as.character(sapply(splitSites, function(x) x[[4]]))

ppe@GeneSymbol <- GeneSymbol
ppe@Residue <- Residue
ppe@Site <- Site
ppe@Sequence <- Sequence

# Alternatively, we can create PhosphoExperiment object as following
ppe <- PhosR::PhosphoExperiment(assays = list(Quantification = as.matrix(PXD001792_raw_hepatocyte)), Site = Site, GeneSymbol = GeneSymbol, Residue = Residue, Sequence = Sequence)

# Lastly add colData information
sample_name = strsplit(gsub("^Intensity.", "", colnames(ppe)), "_")

colData(ppe) = S4Vectors::DataFrame(
    cellline = sapply(sample_name, "[[", 1),
    condition = sapply(sample_name, "[[", 2),
    replicate = sapply(sample_name, "[[", 3)
)

# Have a quick glance of the object
ppe
dim(ppe)
```

# Step 3: Data pre-processing
```{r}
# Log transformation
logmat = log2(assay(ppe, "Quantification"))
logmat[is.infinite(logmat)] = NA
assay(ppe, "Quantification") = logmat

# Filtering using function `selectGrps`
grps <- paste0(colData(ppe)$cellline, "_", colData(ppe)$condition)
ppe <- selectGrps(ppe, grps, 0.5, n=1)
dim(ppe)

# Imputation
set.seed(123)
ppe <- scImpute(ppe, 0.5, grps)
ppe

ppe[,1:6] <- ptImpute(ppe[,7:12], ppe[,1:6], 0.6, 0, paired = FALSE, assay = "imputed")
ppe[,13:18] <- ptImpute(ppe[,19:24], ppe[,13:18], 0.6, 0, paired = FALSE, assay = "imputed")

ppe <- tImpute(ppe, assay = "imputed")
ppe <- medianScaling(ppe, scale = FALSE, assay = "imputed")
ppe

# taking ratio
FL83B.ratio <- assay(ppe, "scaled")[, grep("FL83B_", colnames(ppe))] - 
rowMeans(assay(ppe, "scaled")[,grep("FL83B_Control", colnames(ppe))])
Hepa.ratio <- assay(ppe, "scaled")[, grep("Hepa1.6_", colnames(ppe))] - 
rowMeans(assay(ppe, "scaled")[,grep("Hepa1.6_Control", 
colnames(ppe))])
assay(ppe, "scaled")<- cbind(FL83B.ratio, Hepa.ratio)

assay(ppe, "scaled")
# we will now save this fully processed matrix for use later
PXD001792_ppe_hepatocyte = ppe 
#save(PXD001792_ppe_hepatocyte, file = "../Data/PXD001792_ppe_hepatocyte.RData") ##### check this is the same
```


# Step 4: identifying stably phosphorylated sites
```{r}
# Load datasets
load("../Data/PXD010621_ppe_ESC.RData", verbose = TRUE) #phospho.ESC.ratio.pe
load("../Data/PXD003631_ppe_adipocyte.RData", verbose = TRUE) # phospho.L1.FGF21.ratio.pe
load("../Data/phospho_ppe_adipocyte.RData", verbose = TRUE) # phospho.L1.Ins.ratio.subset.pe.RData

# Simplify names of datasets
ppe1 <- PXD010621_ppe_ESC
ppe2 <- PXD003631_ppe_adipocyte
ppe3 <- phospho_ppe_adipocyte

# Make a list of all PhosphoExperiment objects
ppe.list <- list(ppe1, ppe2, ppe3)

# Make a vector of the selected assays in each of the PhosphoExperiment objects
assays <- rep("Quantification", 3)

# Make a list of grouping information of each dataset
cond.list <- list(
    grp1 = gsub("_.+", "", colnames(ppe1)),
    grp2 = gsub("_r[0-9]", "", colnames(ppe2)),
    grp3 = colnames(ppe3))

# Identifying SPSs by calling getSPS()
inhouse_SPSs <- getSPS(ppe.list, assays = assays, conds = cond.list, residueInfo = TRUE)
```

# Step 4: normalising and batch correction
```{r}
# Load quantification matrix of phosphoproteomic data
load("../Data/PXD019127_ratio_myoblast.RData", verbose = TRUE)
ppe <- PhosphoExperiment(assays = list(Quantification = as.matrix(PXD019127_ratio_myoblast)))

rowNames = strsplit(rownames(ppe), "~")

ppe@GeneSymbol <- toupper(sapply(rowNames, "[[", 2))
ppe@Residue <- gsub("[0-9]","", sapply(rowNames, "[[", 3))
ppe@Site <- as.numeric(gsub("[A-Z]","", sapply(rowNames, "[[", 3)))
ppe@Sequence <- sapply(rowNames, "[[", 4)

# Generate colData
sample_name = strsplit(colnames(ppe), "_")
colData(ppe) = S4Vectors::DataFrame(
    condition = sapply(sample_name, "[[", 1),
    replicate = gsub("exp", "", sapply(sample_name, "[[", 2))
)

# diagnosing batch effect
plotQC(assay(ppe,"Quantification"), panel = "dendrogram", grps = colData(ppe)$condition, labels = colnames(ppe)) + ggplot2::ggtitle("before batch correction")
plotQC(assay(ppe,"Quantification"), panel = "pca", grps = colData(ppe)$condition, labels = colnames(ppe)) + ggplot2::ggtitle("before batch correction")

# correcting batch effect
design = model.matrix(~ colData(ppe)$condition - 1)
head(design) # observe first 6 rows of the design matrix
sites = paste(
    sapply(ppe@GeneSymbol, function(x)x), ";", 
    sapply(ppe@Residue, function(x)x), 
    sapply(ppe@Site, function(x)x), ";", 
    sep = ""
)
data(SPSs)
ctl = which(sites %in% SPSs)

# Run RUV
ppe = RUVphospho(ppe, M = design, k = 3, ctl = ctl)

# QC
# plot after batch correction
p1 = plotQC(assay(ppe,"Quantification"), grps=colData(ppe)$condition, labels = colnames(ppe), panel = "dendrogram")
p2 = plotQC(assay(ppe,"normalised"), grps=colData(ppe)$condition, labels = colnames(ppe), panel="dendrogram")
ggpubr::ggarrange(p1, p2, nrow = 1)

p1 = plotQC(assay(ppe,"Quantification"), panel = "pca", grps=colData(ppe)$condition, labels = colnames(ppe)) +
  ggplot2::ggtitle("Before Batch correction")
p2 = plotQC(assay(ppe,"normalised"), grps=colData(ppe)$condition, labels = colnames(ppe), panel="pca") +
  ggplot2::ggtitle("After Batch correction")
ggpubr::ggarrange(p1, p2, nrow = 2)

# save data
PXD019127_ppe_myoblast = ppe
save(PXD019127_ppe_myoblast, file = "../Data/PXD019127_ppe_myoblast.RData")
```

# Step 5: predicting kinase substrates
```{r}
load("../Data/PXD019127_ppe_myoblast.RData", verbose = TRUE)
ppe = PXD019127_ppe_myoblast
mat = SummarizedExperiment::assay(ppe, "normalised")

mat.mean <- meanAbundance(mat, grps = colData(ppe)$condition)
aov <- matANOVA(mat=mat, grps=colData(ppe)$condition)
idx <- (aov < 0.05) & (rowSums(mat.mean > 0.5) > 0)
mat.reg <- mat[idx, ,drop = FALSE]

mat.std <- PhosR::standardise(mat.reg)
#rownames(mat.std) <- paste0(ppe@GeneSymbol, ";", ppe@Residue, ppe@Site, ";")[idx]
rownames(mat.std) <- sapply(strsplit(rownames(mat.std), "~"), function(x){gsub(" ", "", paste(toupper(x[2]), x[3], "", sep=";"))})

data("KinaseMotifs")
seqs <- ppe@Sequence[idx]
kssMat <- kinaseSubstrateScore(substrate.list = PhosphoSite.mouse, 
                               mat = mat.std, seqs = seqs,
                               numMotif = 5, numSub = 1)
```

```{r}
set.seed(1)
predMat <- kinaseSubstratePred(kssMat, top = 30)
colnames(predMat)
```

```{r}
kinaseOI = c("AKT1")
signalomesRes <- Signalomes(KSR = kssMat, 
                            predMatrix = predMat, 
                            exprsMat = mat.std, 
                            module_res = 5,
                            KOI = kinaseOI)
```

```{r}
### generate palette
my_color_palette <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(8, "Accent"))
kinase_all_color <- my_color_palette(ncol(kssMat$combinedScoreMatrix))
names(kinase_all_color) <- colnames(kssMat$combinedScoreMatrix)
kinase_signalome_color <- kinase_all_color[colnames(predMat)]

plotSignalomeMap(signalomes = signalomesRes, color = kinase_signalome_color)

plotKinaseNetwork(KSR = kssMat, predMatrix = predMat, threshold = 0.95, color = kinase_all_color)
```

